hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-03-04 inst/manuscript/paper.Rmd
one last copy edit
expanding fit_model section

To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature. The function takes two input vectors—one for average incubation temperature (°C) and one for the number of days to a given phenological event—and estimates model parameters (*log~e~a* and *b*) using `stats::nls()`.

This approach allows users to generate models tailored to non-Salmonid species, provided they have experimental or field data linking development to temperature. However, users should be mindful that custom models may not generalize well beyond the temperature range of the data used to fit the model. Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, pending sufficient validation in the literature.

@@

Once uploaded, **hatchR** automatically generates a vizual data check using `plot_check_temp()`. After confirming data integrity, users navigate to the `Model Phenology` window. For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus. The user then chooses multiple spawn dates using an interactive calendar. Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.

>>>>

To expand the applicability of the effective value approach beyond Salmonids, **hatchR** includes a `fit_model()` function, which is species-agnostic as long as development follows a power law relationship with temperature. The function takes two input vectors: average incubation temperature (°C) and the number of days to a given phenological event. A model is then fit to the data using `stats::nls()`, which performs nonlinear least squares regression to estimate the parameters *log~e~a* and *b*. Because the optimization process in `nls()` is sensitive to initial parameter values, `fit_model()` first fits a linear model to the log-transformed data to provide initial parameter estimates, which are then used to fit the nonlinear model.

This approach allows users to generate models tailored to non-Salmonid species, provided they have experimental or field data linking development to temperature. However, users should be mindful of several factors, such as extrapolations risks (models may not generalize beyond the temperature range for which they were parameterized) and species-specific variation (genetic differences among populations may affect developmental responses). Future expansions of **hatchR** could incorporate additional vetted parameterizations for other taxa, such as non-Salmonid fishes, amphibians, invertebrates, provided sufficient validation in the literature.

@@

Once uploaded, **hatchR** automatically generates a visual data check using `plot_check_temp()`. After confirming data integrity, users navigate to the `Model Phenology` window. For this case study, we use the pre-loaded Bull Trout parameterization from @austin2019, selecting the `Existing` model option via dropdown menus. The user then chooses multiple spawn dates using an interactive calendar. Here, we focus on September 30th (in the 2014 spawn year) as outlined in our management scenario.
